Ordinal bhsdtr tutorial

### 2020-02-17 ordinal/ordinal.Rmd
"sample average" mistakes corrected

Now we can translate between the gamma and the criteria vectors (fixed
effects = sample average) for each stimulus:

@@

We will later see that this is our sample average *d'*, but it also
isn't.

@@

If the participant-specific *d'* estimates match, the sample average
*d'* should also match, right? Maybe the SDT model does not fit the
data well?

@@

The discrepancy between the sample average *d'* estimates from both
models is not really a consequence of allowing for negative *d'*
values in the ordinal model. This discrepancy is a consequence of the
assumed distribution of individual shifts / *d'* values. In the
ordinal model we assumed that the shifts (*= d'*) are normally
distributed, whereas in the SDT model we assumed that *log(d')* are
normally distributed. We can see if the assumption that *log(d')* (or
*d'*) are normally distributed is approximately correct in our
dataset. First, let's take a look at participant-specific delta
estimates in the SDT model:

@@

Now we see that the two models give very similar point and interval
estimates of sample-average *d'*.

@@

Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
the *sample-average* pattern of the thresholds seems to be different
for the two stimulus classes (gamma_fixed[.,2] except for
gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
constant across the participants: all the gamma_sd_1[.,2] parameters
seem to be near-zero, except for the gamma_sd_1[4,2] parameter, which
corresponds to the variability in *d'* values. This is weird, since
the pattern of the thresholds for the "noise" stimuli clearly varies
with participants - none of the 95% HPD intervals for the
gamma_sd_1[.,1] parameters include zero: in terms of standard
deviations of posterior distributions of standard deviations (phew)
most are in fact quite far from zero.

@@

If we campared the estimates of the main decision criterion (i.e.,
bias) in SDT models fitted to two different large *samples* of
participants the difference would likely be negligible with very
narrow confidence / credible intervals, even though *individual
participants* clearly differ in how they use the main decision
criterion. By the same token, instability of item parameters in IRT
models cannot be estimated by assuming selective influence and
comparing estimates of "item-parameters" obtained in two different
large samples, because sample average estimates of item-parameters
*hide or misplace* (overdispersion) the intra- or inter-individual
variability in item parameters.

>>>>

Now we can translate between the gamma and the criteria vectors (fixed
effects) for each stimulus:

@@

We will later see that this is our average *d'*, but it also isn't.

@@

If the participant-specific *d'* estimates match, the average *d'*
should also match, right? Maybe the SDT model does not fit the data
well?

@@

The discrepancy between the average *d'* estimates from both models is
not really a consequence of allowing for negative *d'* values in the
ordinal model. This discrepancy is a consequence of the assumed
distribution of individual shifts / *d'* values. In the ordinal model
we assumed that the shifts (*= d'*) are normally distributed, whereas
in the SDT model we assumed that *log(d')* are normally
distributed. We can see if the assumption that *log(d')* (or *d'*) are
normally distributed is approximately correct in our dataset. First,
let's take a look at participant-specific delta estimates in the SDT
model:

@@

Now we see that the two models give very similar point and interval
estimates of average *d'*.

@@

Note the clear pattern for the gamma_sd_1[.,2] parameters. Even though
the *average* pattern of the thresholds seems to be different
for the two stimulus classes (gamma_fixed[.,2] except for
gamma_fixed[4,2] ~ d'), this difference in pattern seems to be fairly
constant across the participants: all the gamma_sd_1[.,2] parameters
seem to be near-zero, except for the gamma_sd_1[4,2] parameter, which
corresponds to the variability in *d'* values. This is weird, since
the pattern of the thresholds for the "noise" stimuli clearly varies
with participants - none of the 95% HPD intervals for the
gamma_sd_1[.,1] parameters include zero: in terms of standard
deviations of posterior distributions of standard deviations (phew)
most are in fact quite far from zero.

@@

If we campared the estimates of the main decision criterion (i.e.,
bias) in SDT models fitted to two different large *samples* of
participants the difference would likely be negligible with very
narrow confidence / credible intervals, even though *individual
participants* clearly differ in how they use the main decision
criterion. By the same token, instability of item parameters in IRT
models cannot be estimated by assuming selective influence and
comparing estimates of "item-parameters" obtained in two different
large samples, because estimates of sample average item-parameters
*hide or misplace* (overdispersion) the intra- or inter-individual
variability in item parameters.
